Grade Retention and Mixed Race

### 2022-08-15 products/paper/main.qmd
Add rough draft of first part of background section

-   The biracial baby boom - why we should care about outcomes for this particular group.

-   What should we expect about the outcomes for biracials relative to their monoracial peer groups?

    -   In each case, our interest is in the outcomes for the biracial group relative to the two constituent groups that their parents identify with. So for example, how to the outcomes of a Black/White individual compare to monoracial Black and monoracial White individuals?
    -   The most direct pathway for understanding these outcomes is via how such individuals are classified by those in positions of power and control, such as teachers, administrators, employers, etc. To the extent that biracial individuals are classified as more like one of their monoracial constituent groups, they will receive the same treatment by individuals and either reap similar benefits or suffer similar discrimination and penalties.
    -   Specifically, I consider two long-standing norms of *hypodescent* and *hyperdescent* in understanding possible placement of biracial individuals (cite F James Davis). Hypodescent indicates a norm in which individuals of mixed racial ancestry are classified as members of the lower status group within a racial hierarchy. In the US, hypodescent is canonically tied to the "one-drop rule", a norm in which those of part-Black ancestry were simply viewed as Black, irrespective of other ancestry (CITATIONS). More generally, if hypodescent is the common practice, we would expect biracial individual's outcomes to be closest those of the monoracial constituent group with the lower outcomes.
    -   Hyperdescent refers to the opposite situation in which a mixed race individual is classified as a member of the higher status group. Because hyperdescent blends individuals into the often unmarked dominant group, it is often more difficult to see. Indigenous ancestry (American Indian, Pacific Islander, and Alaska Native) has historically followed a logic of hyperdescent in which a mixing of indigenous ancestry does not disqualify individuals from claiming membership with another group (most notably Whites). In the case of outcomes, a pattern of hyperdescent would be indicated if biracial individual's outcomes are closest to those of the monoracial constituent group with the higher outcomes.
    -   Discuss work on the contemporary strength of these norms?
    -   The expectations of hypodescent/hyperdescent are that the biracial respondent will have outcomes close to one of the monoracial constituent groups. Another possibility is that biracial respondents will have outcomes intermediate between their monoracial constituent groups. There could be several reasons for this:
        -   The racial ambiguity of biracial respondents may protect them from the worst stigmatization and discrimination in social settings, but by the same coin it may allow them to benefit less from membership in advantaged groups. Thus, a greater recognition of their "blended" ancestry may place them in a halfway position.
        -   Even in cases where hypodescent and hyperdescent are still norms, physical differentation such as skin tone may provide advantages and disadvantages among members who are classified in the same way. We know that skin tone makes a difference even for monoracial Blacks, so the one-drop rule while providing a single classification does not create uniform experiences for all Blacks. (==CITE MONKS WORK HERE I THINK==).
        -   Successful outcomes within social systems depend on a variety of non-material, symbolic resources (e.g. social and cultural capital) that we rarely completely capture in our measurements. Some of the differences in outcomes across racial groups may reflect differences in those resources (CITATIONS?). Given that biracial children have parents belonging to two different races, they may possess such resources that are roughly half-way between the two constituent groups on average, and thus will place them roughly in between the groups. In effect, the in-between outcomes for biracial children may loosely reflect the "black box" package of differences across racial groups that a racial dummy variables stands in for in our research practices.
        -   Two other possibilities don't suggest the actual placement, but rather may modify it in the face of other prevalent patterns. Hybridity itself may be a boon or a burden on biracial individuals that affects their outcomes in, respectively, positive or negative ways.
            -   On the positive side of the ledger, hybridity may enabled a larger cultural "toolkit" that allows mixed race individuals to more easily navigate a variety of social situations. On the negative side of the ledger, the longstanding "marginal man" hypothesis suggests that the dissonance of living in two different worlds may lead to negative mental health outcomes for mixed race individuals, which could spill over into other types of outcomes.

>>>>

How might we expect the outcomes for biracial individuals do differ from their monoracial peers? In every case, the relevant comparison is how a biracial person fares relative to the two constituent groups with which their parents identify. For example, in the case of an individual with one Black and one White parent, we would compare the average outcomes for such individuals to the average outcomes for individuals with two White and two Black parents. The outcome for the Black/White group may be  similar to one of the monoracial groups, it may me somewhere in between the outcomes for the two monoracial groups, or it may be higher or lower than either of the monoracial groups.

The most direct pathway for understanding the placement of this outcome is via how biracial individuals are classified by those in positions of power and control, such as teachers, administrators, and employers. To the extent that biracial individuals are classified as more like one of their monoracial constituent groups. they will receive the same treatment and either reap similar rewards or suffer similar discrimination and penalties.

Two well known norms have long governed such patterns of identification. The pattern of *hypodescent* dictates that individuals with mixed racial ancestry will be classified as members of the lower status group while the pattern of *hyperdescent* dictates that such individuals will be classified as members of the higher status  (==CITATIONS==) Hypodescent, in the form of the "one drop rule" has long characterized the experiences of mixed part-Black individuals who are treated as exclusively Black (==CITE F JAMES DAVIS==). The pattern of hyperdescent blends mixed race individuals into the often unmarked higher status group and is thus more difficult to see in practice. Indigenous ancestry (American Indian, Pacific Islander, and Alaska Native) has historically followed a logic of hyperdescent in which a mixing of indigenous ancestry does not disqualify individuals from claiming to be White (==CITATIONS==). ==NEED TO MIX IN SOME OF THE COMMENTS IN ANN AND ALIYAS RECENT PAPER==

Norms of hyperdescent and hypodescent may account for cases where the biracial group has outcomes similar to one of the monoracial constituent groups. However, two issues may diminish the relevance of these norms for understanding the placement of contemporary biracial populations. First, for individuals with Asian and Latino ancestry, no historical norm governs the placement of individuals with mixed race. Second, the historical norms for those of part-Black and part-Indigenous ancestry may also be breaking down (CITATION). MORE TO DO HERE In both of those cases, it is not entirely clear if such norms exist outside of cases in which the other component in the mixing is Whiteness. Thus, while these norms may give us some historical expectations for certain groups, they are of limited predictive value more generally.

Patterns of hyperdescent and hypodescent expects biracial individuals to have outcomes more similar to one monoracial group. Another possibility is that biracial individuals will have outcomes that are somewhere between their two constituent monoracial groups. Why might such an in-between status exist? There are several possibilities.

Racial ambiguity in physical appearance may help to produce an in-between status. This racial ambiguity may protect biracial respondents from the worst stigmatization and discrimination in social settings. However, by the same coin, it may allow them to benefit less from membership in advantaged groups. The mechanisms for this process may be two-fold. Racial ambiguity may "soften" any particalar experience of discrimination because the person in question does not conform to an ideal type and therefore may be less likely to trigger a sense of racial threat or antipathy. Additionally, racial ambiguity may contribute to more probabilistic and uncertain assessments of racial identification by various social actors that contribute to a lower sum total of disadvantage.

Notably, both of these mechanisms could co-exist with norms of hypo/hyperdescent. We know, for example, that lighter skin tone among monoracial Black individuals is associated with better outcomes (==CITATION==). So, while the one-drop rule provides a consistent classification system, it does not create uniform experiences for all Blacks and skin tone may help to "soften" the experience of discrimination. Similarly, if an in-between status results for biracial individuals becaise they are less consistently classified, that does not mean that the classification system and the norms that underlie it are weakening. (==CITE MONK HERE==)

Aside from ambiguity in physical appearance, a variety of non-material symbolic resources (e.g. social and cultural capital) may also play a role in producing an in-between status. We know that racial disparities persist across a variety of outcomes even after controlling for a wide variety of socioeconomic variables (==CITATIONS==). The residual effect of race here is a "black box" that we often think of as compromising some combination of (a) discrimination and (b) other forms of social and cultural capital not well captured in our data that are differently distributed by race. Biracial individuals may inherit this black box of unobserved resources in weights related to their parent's races, thus helping to produce results that are in-between the two constituent groups. (==I SHOULD GIVE SOME EXAMPLES AND CITATIONS OF WHAT I AM TALKING ABOUT HERE==)

All of these arguments relate to where we might expect the outcomes of biracial individuals to fall relative to their monoracial peers. However, the experience of hybridity itself may have positive and negative effects that do not tell us the location itself but rather may modify it in the context of other patterns. On the positive side, hybridity may enlarge the cultural "toolkit" (==CITATION==) of biracial individuals. This enlarged toolkit may allow such individuals to more easily navigate a variety of social situations (e.g. code switch). On the negative side, Stonequist's (==CITATION==) "marginal man" hypothesis suggests that the dissonance of an identity that crosses strong and salient racial boundaries will be experienced by biracial individuals as not fitting in anywhere. This may result in negative mental health experiences for biracial individuals that can spill over into other types of outcomes.

==NO FORMAL HYPOTHESIS==
